---
title: "sbtools - Exercises"
date: "9999-06-15"
author: "Lindsay R. Carr"
slug: "sbtools-exercises"
image: "usgs-packages/static/img/sbtools.svg"
output: USGSmarkdowntemplates::hugoTraining
parent: Introduction to USGS R Packages
weight: 35
draft: true
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
library(knitr)

knit_hooks$set(plot=function(x, options) {
  sprintf("<img src='../%s%s-%d.%s'/ title='%s'/>", 
          options$fig.path, options$label, options$fig.cur, options$fig.ext, options$fig.cap)

})

opts_chunk$set(
  echo=TRUE,
  fig.path="static/sbtools-exercises/",
  fig.width = 6,
  fig.height = 6,
  fig.cap = "TODO"
)

knit_hooks$set(addToggle = function(before, options, envir) {
    if(before) {
      sprintf('<button class="ToggleButton" onclick="toggle_visibility(\'%1$s\')">Show Answer</button>
              <div id="%1$s" style="display:none">', opts_current$get('label'))
    } else {
      '</div>'
    }
})

set.seed(1)
```

```{r sbtools-auth, echo=FALSE}
# run vizlab::storeSBcreds() once before this can work
home <- path.expand('~')
sbCreds <- file.path(home, ".vizlab/sbCreds")
credList <- readRDS(sbCreds)
un <- rawToChar(credList$username)
pw <- rawToChar(credList$password)
sbtools::authenticate_sb(un, pw)
```

Before starting the exercises, you should make sure that the `sbtools` package is installed and loaded. If you haven't recently updated, you could reinstall the package by running `install.packages('sbtools')` or go to the "Update" button in the "Packages" tab in RStudio.

```{r load-sbtools}
# load the sbtools package
library(sbtools)
```

## Exercise 1

Using querying functions, find out how many ScienceBase items were collected during 2010 related to lake temperature. Hint: use `query_sb` to search using more than one criteria, and use a Lucene query string if you want an exact match. 

```{r addToggle="ON"}
laketemp <- query_sb(query_list = list(q = '', lq = '"lake temperature"', 
                                       dateRange = '{"start":"2010-01-01","end":"2010-12-31"}',
                                       dateType = 'dateCollected'),
                   limit = 100)
length(laketemp)
```

## Exercise 2

Using [this item](https://www.sciencebase.gov/catalog/item/5979248ee4b0ec1a488a49c6), create a new child item. Then, add a subtitle and explanation about this new child item (e.g. "this is a practice item for using sbtools"). Hint: have you authenticated?

```{r addToggle="ON"}
new_item <- item_create(parent_id = "5979248ee4b0ec1a488a49c6", 
                        title = "My new item")
updated_item <- item_update(new_item, 
                            info = list(subTitle = "Test item for sbtools",
                                        body = "This is my practice item for using sbtools."))
```

## Exercise 3

Using the folder "Example NWIS Data" under [this item](https://www.sciencebase.gov/catalog/item/5979248ee4b0ec1a488a49c6), read each file in as a data frame. Hint: JSON files can be read using `fromJSON` from the `jsonlite` package.

```{r addToggle="ON"}
# You can either use the parent folder to determine the "Example NWIS Data" id;
# or go online and get the item id from the end of the URL.

# Use parent folder to get specific child folder reproducibly
all_children <- item_list_children("5979248ee4b0ec1a488a49c6")
children_titles <- unlist(lapply(all_children, function(item) item$title))
folder_item <- all_children[[grep("Example NWIS Data", children_titles)]]

# Use the "Example NWIS Data" id from the website
folder_item <- item_get("59792b28e4b0ec1a488a49dc")

# download the item file info and the files
nwis_files <- item_list_files(folder_item)
nwis_files

ca_dissolvedoxygen <- read.csv(nwis_files$url[1])
id_flow <- read.csv(nwis_files$url[2])
ms_siteinfo <- jsonlite::fromJSON(nwis_files$url[3])
```

```{r cleanup, echo=FALSE, messages=FALSE, warnings=FALSE}
# leave parent folder, but delete anything created in this .Rmd
removed_item <- item_rm(updated_item)
```
