---
title: "Earthquakes and water tables"
author: "Lindsay R. Carr"
date: "October 5, 2017"
output: html_document
---

The following is an example of data analysis workflow borrowed from the blog "Groundwater levels across US affected by Mexico earthquake" by Laura DeCicco (visit that blog post [here](https://owi.usgs.gov/blog/earthquake/)). Note that your plots might differ slightly from the blog post. The blog was created while the data was still provisional and the data may have been adjusted during review.

## First, get data.

```{r get-data, warning=FALSE, message=FALSE}
library(dataRetrieval)

sites <- c("402411077374801", "364818094185302", "405215084335400", "370812080261901",
           "393851077343001", "444302070252401", "324307117063502", "421157075535401",
           "373904118570701", "343457096404501", "401804074432601", "292618099165901")

gw_data <- readNWISuv(sites, 
                      parameterCd = "72019", 
                      startDate = "2017-09-07",
                      endDate = "2017-09-08")

unique_sites <- attr(gw_data, "siteInfo")

```

## Clean, arrange, and prepare data

```{r clean-data, warning=FALSE, message=FALSE}
library(dplyr)

# order site info data from Southern-most sites to Northern-most
south_to_north <- arrange(unique_sites, desc(dec_lat_va))

# subset data so it only includes September 8th, 2017
startDateTime <- as.POSIXct("2017-09-07 00:00:00", tz="UTC")
endDateTime <- as.POSIXct("2017-09-09 00:00:00", tz="UTC")
gw_data_sep8 <- gw_data %>%
  filter(dateTime > startDateTime,
         dateTime < endDateTime) 

# use the dataRetrieval function to get more user-friendly column names
# WLBLS = water level below surface
gw_data_sep8 <- renameNWISColumns(gw_data_sep8)
```

## Plot data

```{r plot-data, warning=FALSE, message=FALSE, fig.height=3, fig.width=10}
library(ggplot2)

# get/set ylabel & xlabel from data
y_label <- attr(gw_data_sep8, "variableInfo")$variableName
x_label <- "UTC time"

# gather colors to be used for each site
col_values <- c("#DCDA4B","#999999","#00FFFF","#CEA226","#CC79A7","#4E26CE",
                "#0000ff","#78C15A","#79AEAE","#FF0000","#00FF00","#B1611D",
                "#FFA500","#F4426e", "#800000", "#808000")

n_sites <- nrow(south_to_north)

# loop through and make a plot for each of the 12 sites
for(i in 1:n_sites){
  
  site_i <- south_to_north[i, "site_no"]
  gw_current_site <- filter(gw_data_sep8, site_no == site_i)
  
  plot_current <- ggplot(data = gw_current_site) +
    geom_line(aes(x = dateTime, y = WLBLS_Inst), color = col_values[i], size=1.5) +
    theme_bw() +
    scale_y_continuous(trans = "reverse") +
    ylab(label = y_label) +
    xlab(label = x_label) +
    ggtitle(paste("Site", site_i)) +
    theme(strip.background = element_blank(),
          strip.text.y = element_text(angle = 0),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank())
  
  print(plot_current)
}
```

We show a `for` loop here because that is what we will learn in this class. `ggplot2` has another feature that allows us to get all of these unique site timeseries plots with less code using the `facet` function. 

```{r ggplot-data, warning=FALSE, message=FALSE, fig.height=8, fig.width=10}
# ggplot2 faceting

# create timeseries plot for each site w/ shared x-axis
ggplot(data = gw_data_sep8) +
  geom_line(aes(x = dateTime, y = WLBLS_Inst, col = site_no)) +
  theme_bw() +
  scale_y_continuous(trans = "reverse") +
  scale_color_manual(values = col_values) +
  facet_grid(site_no ~ ., scales = "free") +
  ylab(label = y_label) +
  xlab(label = x_label) +
  theme(strip.background = element_blank(),
        strip.text.y = element_text(angle = 0),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position="none",
        panel.spacing.y=unit(0.05, "lines"))

```

## Spice it up with a map

```{r map-sites, warning=FALSE, message=FALSE, fig.height=6}
library(maps)

map("state", fill = TRUE, col = "grey90", border = "white")
points(south_to_north$dec_lon_va, south_to_north$dec_lat_va, pch=20, col = col_values)

```
